Condensed Matter Field Theory notes

### 2017-03-16 Condensed Matter Field Theory note/CMF-Atland.tex
[minor] notes, especially about Grassmann calculus.

Use $aa^\dagger = 1+ a^\dagger a$. And observe:
\begin{align*}
    a(a^\dagger)^n &= (a^\dagger)^{n-1}+ a^\dagger a (a^\dagger)^{n-1} \\
    &= (a^\dagger)^{n-1}+ (a^\dagger)^{n-1}+ (a^\dagger)^2 a (a^\dagger)^{n-2}
    \\
    &= 3(a^\dagger)^{n-1} + (a^\dagger)^3 a (a^\dagger)^{n-3} \\
    &\cdots \\
    &= n (a^\dagger)^{n-1} + (a^\dagger)^{n} a
\end{align*}
Then (notice $a\ket{0}=0$):
\begin{align*}
    \sum_n a \frac{\phi^n (a^\dagger)^n}{n!}\ket{0} 
    &= \sum_n \phi\, \phi^{n-1} \frac{n(a^\dagger)^{n-1}+(a^\dagger)^n a}{n!}\ket{0}\\
    &= \sum_n \phi\, \phi^{n-1} \frac{n(a^\dagger)^{n-1}}{n!}\ket{0} \\
    &= \phi\sum_n \phi^{n} \frac{(a^\dagger)^{n}}{n!}\ket{0} \\
    &= \phi\exp{\phi a^\dagger}\ket{0}
\end{align*}
\section{License}
The entire content of this work (including the source code
for TeX files and the generated PDF documents) by 
Hongxiang Chen (nicknamed we.taper, or just Taper) is
licensed under a 
\href{http://creativecommons.org/licenses/by-nc-sa/4.0/}{Creative 
Commons Attribution-NonCommercial-ShareAlike 4.0 International 
License}. Permissions beyond the scope of this 
license may be available at \url{mailto:we.taper[at]gmail[dot]com}.
\end{document}

>>>>

Use $aa^\dagger = 1+ a^\dagger a$. And observe:
\begin{align*}
    a(a^\dagger)^n &= (a^\dagger)^{n-1}+ a^\dagger a (a^\dagger)^{n-1} \\
    &= (a^\dagger)^{n-1}+ (a^\dagger)^{n-1}+ (a^\dagger)^2 a (a^\dagger)^{n-2}
    \\
    &= 3(a^\dagger)^{n-1} + (a^\dagger)^3 a (a^\dagger)^{n-3} \\
    &\cdots \\
    &= n (a^\dagger)^{n-1} + (a^\dagger)^{n} a
\end{align*}
Then (notice $a\ket{0}=0$):
\begin{align*}
    \sum_n a \frac{\phi^n (a^\dagger)^n}{n!}\ket{0} 
    &= \sum_n \phi\, \phi^{n-1} \frac{n(a^\dagger)^{n-1}+(a^\dagger)^n a}{n!}\ket{0}\\
    &= \sum_n \phi\, \phi^{n-1} \frac{n(a^\dagger)^{n-1}}{n!}\ket{0} \\
    &= \phi\sum_n \phi^{n} \frac{(a^\dagger)^{n}}{n!}\ket{0} \\
    &= \phi\exp{\phi a^\dagger}\ket{0}
\end{align*}
\section{Something about Grassmann Algebra, Grassmann Calculus}

I believe that there are several important points missing in the book, when he
talks about Grassmann.

\paragraph{The first} thing is the derivative rule. The differentiation with respect to
Grassmann numbers should follow some version of \nomen{graded Leibniz rule}. But
I am not sure about the details for the following reason. Consider the
traditional graded Leibniz rule:
\begin{equation}
    \dd (fg) = (\dd f) g + (-1)^{[f]} f\dd g
\end{equation}
where $[f]$ is something defined as the parity, or the degree of f.

Now comes the problem: how to define a grade for something like:
\begin{align}
    f &= \eta_1 +\eta_2 + \eta_3 + \eta_1\eta_2 +\eta_1\eta_2\eta_3 \\
    g &= \eta_1 -\eta_2 -\eta_3 + \eta_1\eta_3 - \eta_1\eta_2\eta_3
\end{align}

\todo{understand graded Leibniz rule}

By calculation, I found:
\begin{align}
    \frac{\partial }{\partial \eta_1}(fg) &= -2\eta_2-2\eta_3-2\eta_2\eta_3 \\
    (\frac{\partial }{\partial \eta_1}f)g &=
        \eta_1-\eta_2-\eta_3+\eta_1\eta_3-\eta_1\eta_2-\eta_2\eta_3-\eta_1\eta_2\eta_3
    \\
    f( \frac{\partial }{\partial \eta_1}g) &=
        \eta_1+\eta_2+\eta_3+ \eta_1\eta_3 +\eta_1\eta_2+\eta_2\eta_3
        +\eta_1\eta_2\eta_3
\end{align}
It is easy to see that:
\begin{equation}
    \pdv{\eta_1}(fg) \neq (\pdv{\eta_1}f)g \pm f(\pdv{\eta_1} g)
\end{equation}

A more general calculation gives:
\begin{align}
    \pdv{\eta}\left[ (a+\eta b)(c+\eta d) \right] &= bc + (-1)^{[a]} ad \\
    \left(\pdv{\eta}(a+\eta b)\right) (c+\eta d) &= bc + b\eta d \\
    (a+\eta b)\left(\pdv{\eta}(c+\eta d)\right) &= ad + \eta bd
\end{align}
where $(-1)^{[a]}$ is the number such that $a\eta = (-1)^{[a]}\eta a$. So, there
is no way to reconcile the above calculations into something like:
\begin{align*}
    \pdv{\eta}\left[ (a+\eta b)(c+\eta d) \right] = 
    \left(\pdv{\eta}(a+\eta b)\right) (c+\eta d)  
    + (-1)^?  (a+\eta b)\left(\pdv{\eta}(c+\eta d)\right)
\end{align*}

\paragraph{The second} point is very important. It is about the postulations of
Grassmann variables. I postulate that (in addition to the postulations in p.162
of \cite{Altland2010}):
\begin{enumerate}
    \item All $\eta_i,\bar{\eta}_i,a_i,a^\dagger_i$ anti-commute with each
        other. From this we also have $\eta_i^2=0$, etc.
    \item $\{\dd \eta_i,\dd\eta_j\}=0$ for ($i\neq j$).
    \item $\int \dd{\eta} (af(\eta) + bg(\eta) )
        = a\int\dd{\eta}f(\eta) + b\int\dd{\eta}g(\eta) $
    \item $\{\eta_i,\dd{\eta_j}\}=0$, for $i\neq j$.
    \item $\{\bar{\eta}_i,\dd{\eta_j}\}=0$, for $\forall i,j$.
    \item $\pdv{\eta} (af(\eta)) = a\pdv{\eta}f(\eta)$, for $a\in\C$.
\end{enumerate}
It is not so simple to tell whether we should have 
$$[\eta_i,\dd\eta_i]=0 \,\text{?}$$ or
$$\{\eta_i,\dd\eta_i\}=0\,\text{?}$$
But we should have 
$$\{\eta_i,\dd{\eta_j}\}=0 \quad\text{for $i\neq j$} $$
For if $[\eta_i,\dd{\eta_j}]=0$, then we have a contradiction:
\begin{align}
    \int\dd{\eta_j} \eta_i\eta_j &= \eta_i \int\dd{\eta_j}\eta_j = \eta_i
    \nonumber\\
    &= -\int\dd{\eta_j}\eta_j \eta_i = -\eta_i\quad \text{contradiction!}
\end{align}
And similar argument applies for $\{\bar{\eta}_i,\dd{\eta_j}\}=0$, $\forall
i,j$.

There is also another point missing. The relations between $\eta$ and
$\ket{0}\bra{0}$ is ambiguous. It seems that there is no harm in assuming:
\begin{equation}
    \eta \ket{0}\bra{0} = \ket{0}\bra{0}\eta
\end{equation}
Because, if so, we have:
\begin{equation}
    a\ket{\eta}\bra{\eta} = \eta\ket{\eta}\bra{\eta} = \ket{\eta}\bra{\eta}\eta
\end{equation}
without conflict with:
\begin{equation}
    \bar{\eta}\ket{\eta}\bra{\eta} = \ket{\eta}\bra{\eta}\bar{\eta} =
    \ket{\eta}\bra{\eta} a^\dagger
\end{equation}

With above assumptions, I have:
\begin{thm}
    Assuming that $a,b,c,d$ either commute or anticommute with $\eta$.
    \begin{equation}
        \int\dd{\eta} \left(\pdv{\eta}(a+\eta b) \right)(c+\eta d)
        = (-1)^s\int\dd{\eta} (a+\eta b) \left(\pdv{\eta}(c+\eta d)\right)
        \label{eq:inteByPart-grass}
    \end{equation}
    where $s$ is such that $b\eta = (-1)^s \eta b$.
\end{thm}
This can also be generalized into
\begin{equation}
    \int\dd{\eta} \left(\pdv{\eta}f(\eta) \right) g(\eta)
    = (-1)^s\int\dd{\eta} f(\eta) \left(\pdv{\eta}g(\eta)\right)
\end{equation}
where $s$ is some appropriate constant as in eq~\ref{eq:inteByPart-grass}. Here
we have assumed that any $f(\eta)$, we can decompose it into $f=a+\eta b$, where
$a$ and $b$ do not contain $\eta$. We also assume $a,b$ either commute or
anticommute with $\eta$.
\begin{proof}
    Direct calculation.
\end{proof}

\paragraph{Change of variables in integration}
First we consider a simple example.
\begin{equation}
    \int \dd{\phi} a+b\phi = b
\end{equation}
Let $v = \lambda \phi$, since
\begin{equation}
    \int\dd{v} a+ b\frac{v}{\lambda} = \frac{b}{\lambda}
\end{equation}
We naturally require
\begin{equation}
    \dd{\phi} = \lambda \dd{v}
\end{equation}
For multivariable case, it is easy to guess that:
\begin{align}
    \vb{v} &= M\vb{\phi} \\
    \dd{\phi_1}\dd{\phi_2}\cdots \dd{\phi_N} &= \det(M)\dd{v_1}\dd{v_2}\cdots\dd{v_N}
\end{align}
(Took from
\href{https://en.wikipedia.org/wiki/Berezin_integral#Change_of_Grassmann_variables}{Wikipedia
Berezin Integral}.)

\section{License}
The entire content of this work (including the source code
for TeX files and the generated PDF documents) by 
Hongxiang Chen (nicknamed we.taper, or just Taper) is
licensed under a 
\href{http://creativecommons.org/licenses/by-nc-sa/4.0/}{Creative 
Commons Attribution-NonCommercial-ShareAlike 4.0 International 
License}. Permissions beyond the scope of this 
license may be available at \url{mailto:we.taper[at]gmail[dot]com}.
\end{document}
